---
title: "GDP Modeling"
output: html_notebook
---

```{r setup}
library('eiamod')
library('dplyr', warn.conflicts = FALSE)
library('tibble')
library('ggplot2')
library('ggthemes')
library('earth')
```

## Model I: Total temperature-adjusted energy vs. GDP

The first principal component over the spatial component of the data is basically a modestly
weighted sum over the regions.  For simplicity, we can just approximate this as a sum over
the regions.  It's less clear what to do with the time dependence.  As a first attempt, we'll
just sum over that too and see what we get.

```{r model1.data}
elec <- group_by(residual_load, EconYear, quarter) %>% summarise(elec=sum(residual_load))
energy <- group_by(energy_weekly, EconYear, quarter) %>% summarise(petrol=sum(petrol), gas=sum(gas))
gdp <- select(gdp_quarterly, EconYear, quarter, gdp)
m1data <- full_join(elec, energy, by=c('EconYear', 'quarter')) %>% 
    full_join(gdp, by=c('EconYear', 'quarter'))
```

```{r model1.lm}
m0lm <- lm(gdp ~ elec, m1data)
m0pglm <- lm(gdp ~ petrol+gas, m1data)
m1lm <- lm(gdp ~ elec+petrol+gas, m1data)
summary(m1lm)
gdp1lm <- predict(m1lm)
plot(m1lm)
plot(m1data$gdp, gdp1lm, xlab='GDP (observed)', ylab='GDP (predicted)')
```

```{r model1.earth}
m1ea <- earth(gdp ~ elec+petrol+gas, m1data)
summary(m1ea)
gdp1ea <- predict(m1ea)
plot(m1ea)
plot(m1data$gdp, gdp1ea, xlab='GDP (observed)', ylab='GDP (predicted)')
```

## Model II:  Timewise PCA of total temperature-adjusted electricity plus other energy vs. GDP

```{r model2.elecdata}
elec <- group_by(residual_load, EconYear, quarter, week, time) %>% 
    summarise(elec=sum(residual_load)) %>% ungroup() %>%
    arrange(EconYear, quarter, week, time)
elecdata_weekly <- matrix(elec$elec, ncol=168)     # 168 hours in a week
pcdata <- prcomp(elecdata_weekly, scale.=TRUE, retx=TRUE)
rpwr <- pcdata$sdev^2/sum(pcdata$sdev^2)
```

```{r model2.pcplots}
rp <- rpwr[1:25]
ggplot(mapping=aes(x=seq_along(rp), y=rp)) + geom_point(color='lightgrey') + 
    xlab('index') + ylab('relative power') +
    theme_solarized_2(light=FALSE) + scale_color_solarized()
pcpltdata <- as.data.frame(pcdata$rotation[,1:6]) %>% mutate(hour=seq(1,168)) %>%
    tidyr::gather('pc','value', -hour)
ggplot(data=pcpltdata, aes(x=hour, y=value, color=pc)) + geom_line() + facet_wrap(~pc) +
    scale_x_continuous(breaks=24*seq(0,7), minor_breaks=(24*seq(1,7))-12) +
    theme_solarized_2(light=FALSE) + scale_color_solarized()
```

It appears that there are about 6 components with significant power, though maybe one could
argue for 8 or 9, with a sufficiently generous definition of "significant".  We will continue
to focus on just the first 6, which we have plotted above.  Evidently each of the components
is responding to periodicity on one or more scales.  Note, however, that each hour was 
independently centered and scaled for the PCA, so this is not a response to periodic behavior
in the raw load, but rather to periodic deviations from the mean load (across the entire dataset) 
for the hour.

To get a better understanding of the nature of these periodic fluctuations, we can do a power 
spectrum analysis.

```{r model2.pcps}
pc16 <- pcdata$rotation[,1:6]
pcmeans <- colMeans(pc16)
pcctr <- sweep(pc16, 2, pcmeans)
nsamp <- nrow(pc16)
inq <- 1 + nsamp/2
pcspec <- Mod(mvfft(pcctr))[1:inq,]
f <- seq(0,inq-1) / nsamp
tau <- 1/f   # tau in days
pltpwrspec <- as.data.frame(pcspec) %>% mutate(f=f, tau=tau) %>% tidyr::gather('pc','sqrtpwr', -f, -tau)
brkhr <- c(48, 24, 12, 6, 3)
brkf <- 1/brkhr
brkmin <- 1/seq(1,12)
ggplot(data=pltpwrspec, aes(x=f, y=sqrtpwr, fill=pc)) + 
    #geom_smooth(se=FALSE, span=0.1) + 
    geom_col() +
    facet_wrap(~pc) +
    scale_x_continuous(name='tau (hr)', breaks=brkf, labels=as.character(brkhr), minor_breaks=brkmin) +
    theme_solarized_2(light=FALSE) + scale_fill_solarized()
```

```{r model2.lm}
## Get a table of weeks
weektbl <- count(elec, EconYear, quarter, week) %>% arrange(EconYear, quarter, week)
invisible(assertthat::assert_that(all(weektbl$n == 168)))
invisible(assertthat::assert_that(nrow(weektbl) == nrow(pcdata$x)))

## Now, the matrix pcdata$x has one row for each week; PC projection coefficients are in 
## columns.  We'll keep 1-6.  For each quarter, sum up the statistics over weeks.
m2data <- select(weektbl, -n) %>% 
    bind_cols(as.data.frame(pcdata$x[,1:6])) %>%
    group_by(EconYear, quarter) %>% summarise(PC1=sum(PC1), PC2=sum(PC2), PC3=sum(PC3),
                                              PC4=sum(PC4), PC5=sum(PC5), PC6=sum(PC6)) %>%
    ungroup() %>%
    full_join(energy, by=c('EconYear', 'quarter')) %>%
    full_join(gdp, by=c('EconYear', 'quarter')) 

m2lm <- list()
length(m2lm) <- 6
m2lm[[1]] <- lm(gdp ~ PC1+petrol+gas, m2data)
m2lm[[2]] <- lm(gdp ~ PC1+PC2+petrol+gas, m2data)
m2lm[[3]] <- lm(gdp ~ PC1+PC2+PC3+petrol+gas, m2data)
m2lm[[4]] <- lm(gdp ~ PC1+PC2+PC3+PC4+petrol+gas, m2data)
m2lm[[5]] <- lm(gdp ~ PC1+PC2+PC3+PC4+PC5+petrol+gas, m2data)
m2lm[[6]] <- lm(gdp ~ PC1+PC2+PC3+PC4+PC5+PC6+petrol+gas, m2data)

m2lmstats <- bind_rows(lapply(m2lm, broom::glance))
m2lmstats
```
