---
title: "Simple models for load from temperature"
output: 
  html_notebook: 
    fig_width: 6.5
    highlight: pygments
    theme: readable
---


```{r setup}
library(eiamod)
library(earth)
#library(randomForest)

rsltsummary <- function(rslt, stat='r.squared') {
    dplyr::bind_rows(
        lapply(names(rslt), function(rgn) {
            r2 <- summary(rslt[[rgn]])[[stat]]
            data.frame(rgn=rgn, r2=r2, stringsAsFactors = FALSE)   
        })
    )
}

DO_SAVE_DATA <- FALSE
```

```{r stdata}
stdata <- standardize_byrgn(region_hourly, c('load','temperature'))
```

## Linear model

Run linear models by region as a baseline for performance.
```{r lm}
rsltlm <- load_model(hourly_data=stdata)
rsltsummary(rsltlm)
```

We expect an increase in load for both high and low temperatures, so we can retry the
linear model with a squared term.
```{r lmsqr}
rsltlmsqr <- load_model(hourly_data=stdata, formula=load~temperature + I(temperature^2))
rsltsummary(rsltlmsqr)
```

## Fits using MARS models
```{r earth}
rsltearth <- load_model(earth, hourly_data=stdata)
rsltsummary(rsltearth, 'grsq')
```
Interesting.  These compare pretty favorably to the linear model using the squared term.  It looks
like it's a little higher across the board.  We can try adding more terms to see if we get anything 
new.
```{r earthd2}
rsltearthd2 <- load_model(earth, stdata, formula=load~temperature+I(temperature^2), degree=3)
rsltsummary(rsltearth, 'grsq')
```
Same result as the above.  I would interpret this as saying that the main thing here is
allowing the model to increase in both directions, low and high.  The $T$^2 term does this,
and it also adds some curvature to the relation, but the curvature isn't doing anythng for us
because the earth model with just temperature does just as well.

## Random forest models
I don't expect this to add a whole lot of value, since in the end we have only the one predictor,
but it's worth seeing what we can get with an off the shelf ML algorithm.  (Removed because it 
takes a long time to run and doesn't perform very well.)
```{r randomforest}
##rsltrf <- load_model(randomForest, stdata, ntree=100)
##rsltsummary(rsltrf, 'rsq')
```

## Residual load dataset

The MARS model seems to perform the best, so we will use that one to take out the temperature
component of the load.

```{r loadxtemp}
ep <- as.vector(sapply(rsltearth, predict))
plot(stdata$load, ep, pch='.')
stdata <- dplyr::mutate(stdata, residual_load = load - ep)
hist(stdata$residual_load, breaks=100)
```

This data is going to be the basis for our modeling, so we will store this as
`residual_load` in the package data.

```{r savedata}
if(DO_SAVE_DATA) {
    residual_load <- dplyr::select(stdata, EconYear, quarter, week, region, time, residual_load)
    usethis::use_data(residual_load, compress='xz')
}
```
